---
title: "Sepsis and Lactate Health Disparity"
author: "Justin Quion, Anamik Jhunjhunwala, Andrew Bonney, Pratiksha Pradhan, Neel Vahil, Suresh Agarwal, Teresa A. Rincon"
date: "2023-08-05"
output:
  word_document: default
  html_document: default
---


```{r setup, include=FALSE}
library(scales)
library(tidyr)
library(dplyr)
library(knitr)
require(ggthemes)
library(ggplot2)
library(tidyverse)  
library(gridExtra)
library("scatterplot3d")
library(outliers)
library(reshape2)
library(lubridate)
library(plotly)
library(qcc)
library(rpart)
library(randomForest)
library(rpart.plot)
library("outliers")
library(EnvStats)
library(car)
library(tidyverse)
library(rpart)
library(randomForest)
library(rsample)
library(recipes)
library(car)
library(censReg)
library(caret)
library(syuzhet)
library(readxl)
library(sentimentr)
library(tidytext)
library(topicmodels)
library(tm)
library(MASS)
library(survival)
library(DataExplorer)
library(plyr)                                                 
library(readr)
library(data.table)
# library(regclass)
# library(FrF2)
```


# Part I - Importing and cleaning the data

```{r}
# Importing the data
mydata <- read_csv("SCCM_cohort.csv")
subset_data <- mydata[complete.cases(mydata[["lactate_freq_day1"]]), ]
head(subset_data)
summary(subset_data)
```


```{r}
#Checking the categorical variables to make sure that they are consistent
unique(subset_data$gender)
unique(subset_data$race_group)
unique(subset_data$language)
unique(subset_data$insurance)
```


# Part II - Exploratory Data Analysis


Single variable data exploration -- 
```{r}
#Histogram and density plot of Charges
ggplot(data = subset_data, aes(x = admission_age)) + 
geom_histogram(aes(y=..density..), colour="black", fill="white") + 
stat_density(alpha=.2, fill="#FF0000") + 
geom_vline(aes(xintercept=mean(admission_age)), color="blue", linetype="dashed", size=1)

#Histogram and density plot of bmi
par(mfrow=c(1,2))
ggplot(data = subset_data, aes(x = weight_admit)) + 
geom_histogram(aes(y=..density..), colour="black", fill="white") + 
stat_density(alpha=.2, fill="#0000FF")+ 
geom_vline(aes(xintercept=mean(weight_admit)), color="blue", linetype="dashed", size=1)


#Histogram and density plot of age
par(mfrow=c(1,2))
ggplot(data = subset_data, aes(x = charlson_comorbidity_index)) + 
geom_histogram(aes(y=..density..), colour="black", fill="white") + 
stat_density(alpha=.2, fill="#FFFF00")+ 
geom_vline(aes(xintercept=mean(charlson_comorbidity_index)), color="blue", linetype="dashed", size=1)

#Histogram and density plot of children
par(mfrow=c(1,2))
ggplot(data = subset_data, aes(x = SOFA)) + 
geom_histogram(aes(y=..density..), colour="black", fill="white") + 
stat_density(alpha=.2, fill="#556B2F")+ 
geom_vline(aes(xintercept=mean(SOFA)), color="blue", linetype="dashed", size=1)
```


```{r}

# Assuming you have a dataframe named 'data'
column_name <- "lactate_freq_day1" # Name of the column with missing values

# Subset the dataframe to include only rows with missing values in the specified column
missing_rows <- mydata[is.na(mydata$lactate_freq_day1),]
value_counts <- table(missing_rows [["race_group"]])
value_counts2 <- table(mydata[["race_group"]])
print(value_counts)
print(value_counts2)

percent_VC <- value_counts/value_counts2
print(percent_VC)

categories <- unique(subset_data$race_group)
values <- percent_VC
barplot(values, names.arg = categories, xlab = "Categories", ylab = "Values", main = "Simple Bar Graph")

# Create two example vectors
vector1 <- percent_VC
vector2 <- subset_data$race_group

# Create a dataframe with two vectors
my_dataframe <- as.data.frame(percent_VC)


ggplot(my_dataframe, aes(x = Var1, y = Freq, fill = categories)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +
  labs(x = "Racial Background", y = "Lactate Measurements not performed") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  ) +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62", "#8da0cb",  "#390a28", "#49b4da"))
```





Univariate data exploration -- 
```{r}
ggplot(subset_data, aes(x=gender, y = lactate_freq_day1, color=gender)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))


ggplot(subset_data, aes(x=race_group, y=lactate_freq_day1, color=race_group)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(subset_data, aes(x=language, y=lactate_freq_day1, color=language)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(subset_data, aes(x=insurance, y=lactate_freq_day1, color=insurance)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

```



Multivaraite data exploration
By gender
```{r}
# ggplot(mydata, aes(x=gender, y=lactate_freq_day1, , color=gender)) + 
#   geom_boxplot(notch=TRUE) +
#   stat_summary(fun=mean, geom="point", shape=23, size=4) + 
#   geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=race_group, y=lactate_freq_day1, , color=gender)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=language, y=lactate_freq_day1, , color=gender)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=insurance, y=lactate_freq_day1, , color=gender)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

```

Multivariate data exploration
By race group
```{r}
ggplot(mydata, aes(x=gender, y=lactate_freq_day1, , color=race_group)) +
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) +
  geom_jitter(shape=16, position=position_jitter(0.2))

# ggplot(mydata, aes(x=race_group, y=lactate_freq_day1, , color=race group)) + 
#   geom_boxplot(notch=TRUE) +
#   stat_summary(fun=mean, geom="point", shape=23, size=4) + 
#   geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=language, y=lactate_freq_day1, , color=race_group)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=insurance, y=lactate_freq_day1, , color=race_group)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

```


Multivaraite data exploration
By language
```{r}
ggplot(mydata, aes(x=gender, y=lactate_freq_day1, , color=language)) +
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) +
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=race_group, y=lactate_freq_day1, , color=language)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

# ggplot(mydata, aes(x=language, y=lactate_freq_day1, , color=language)) + 
#   geom_boxplot(notch=TRUE) +
#   stat_summary(fun=mean, geom="point", shape=23, size=4) + 
#   geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=insurance, y=lactate_freq_day1, , color=language)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

```

Multivaraite data exploration
By insurance
```{r}
ggplot(mydata, aes(x=gender, y=lactate_freq_day1, , color=insurance)) +
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) +
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=race_group, y=lactate_freq_day1, , color=insurance)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

ggplot(mydata, aes(x=language, y=lactate_freq_day1, , color=insurance)) + 
  geom_boxplot(notch=TRUE) +
  stat_summary(fun=mean, geom="point", shape=23, size=4) + 
  geom_jitter(shape=16, position=position_jitter(0.2))

# ggplot(mydata, aes(x=insurance, y=lactate_freq_day1, , color=insurance)) + 
#   geom_boxplot(notch=TRUE) +
#   stat_summary(fun=mean, geom="point", shape=23, size=4) + 
#   geom_jitter(shape=16, position=position_jitter(0.2))

```

Multivariate data exploration
By race group
```{r}
ggplot(subset_data, aes(x=admission_age, y=lactate_freq_day1, color=race_group, shape=race_group)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=weight_admit, y=lactate_freq_day1, color=race_group, shape=race_group)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=prob_mort, y=lactate_freq_day1, color=race_group, shape=race_group)) +
  geom_smooth(method=lm) + 
  geom_density_2d()
```


Multivariate data exploration
By language
```{r}
ggplot(subset_data, aes(x=admission_age, y=lactate_freq_day1, color=language, shape=language)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=weight_admit, y=lactate_freq_day1, color=language, shape=language)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=prob_mort, y=lactate_freq_day1, color=language, shape=language)) +
  geom_smooth(method=lm) + 
  geom_density_2d()
```


Multivariate data exploration
By gender
```{r}
ggplot(subset_data, aes(x=admission_age, y=lactate_freq_day1, color=gender, shape=gender)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=weight_admit, y=lactate_freq_day1, color=gender, shape=gender)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=prob_mort, y=lactate_freq_day1, color=gender, shape=gender)) +
  geom_smooth(method=lm) + 
  geom_density_2d()
```


Multivariate data exploration
By insurance
```{r}
ggplot(subset_data, aes(x=admission_age, y=lactate_freq_day1, color=insurance, shape=insurance)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=weight_admit, y=lactate_freq_day1, color=insurance, shape=insurance)) +
  geom_smooth(method=lm) + 
  geom_density_2d()

ggplot(subset_data, aes(x=prob_mort, y=lactate_freq_day1, color=insurance, shape=insurance)) +
  geom_smooth(method=lm) + 
  geom_density_2d()
```


# Questions to explore

# Part III - Building the Linear regression model
```{r}
model <- lm(lactate_freq_day1 ~ admission_age + gender + race_group + language + weight_admit + insurance + SOFA + charlson_comorbidity_index, data = subset_data)
summary(model)
sigma(model)

plot(model)
```
 #Trying an Elastic Net Linear Regression Model
 
```{r}
# Assuming you have a dataframe named 'data'
selected_columns_A <- c("admission_age","gender","race_group","language","weight_admit","insurance","SOFA","charlson_comorbidity_index")  # Names of columns to convert

selected_columns_B <- c("admission_age","gender","race_group","language","weight_admit","insurance","adm_type","SOFA","charlson_comorbidity_index")  # Names of columns to convert

selected_columns_C <- c("admission_age","gender","race_group","language","weight_admit","insurance","adm_type","SOFA","charlson_comorbidity_index")  # Names of columns to convert

selected_columns_D <- c("admission_age","gender","race_group","language","weight_admit","insurance","adm_type","SOFA","charlson_comorbidity_index")  # Names of columns to convert


```
 

# ```{r}
# model <- lm(lactate_freq_day1 ~ admission_age + gender + race + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index, data = mydata)
# summary(model)
# sigma(model)
# ```


```{r}
subset_data <- mydata[complete.cases(mydata[["lactate_freq_day1"]]), ]
selected_data <- subset_data[selected_columns]

selected_data <- selected_data %>%
  mutate(across(c("admission_age","gender","race_group","language","weight_admit","insurance","SOFA","charlson_comorbidity_index")), ~ as.factor(.)) %>%
  mutate(across(c("admission_age","gender","race_group","language","weight_admit","insurance","SOFA","charlson_comorbidity_index"), ~ as.numeric(.)))  # Convert to numeric (0 or 1)
```

```{r}
library(dplyr)
library(glmnet)

# Convert selected columns to a matrix
X <- selected_data %>%
  as.matrix()

y <- subset_data$lactate_freq_day1 %>%
  as.matrix()

y <- as.matrix(subset_data$lactate_freq_day1)
elastic_net_model <- cv.glmnet(selected_matrix, y , alpha = 0.5)  # 0.5 represents a balance between L1 and L2 penalties

# Print the model summary
print(elastic_net_model)
```


```{r}
model <- lm(lactate_day1 ~ admission_age + gender + race_group + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index, data = mydata)
summary(model)
sigma(model)
```



```{r}
model <- lm(lactate_freq_day1 ~ admission_age + gender + race_group + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)

# resp_rate_mean	mbp_mean	heart_rate_mean	temperature_mean	spo2_mean

```





```{r}
model <- lm(lactate_freq_day1 ~ weight_admit + insurance + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)

# resp_rate_mean	mbp_mean	heart_rate_mean	temperature_mean	spo2_mean

```



```{r}
model <- lm(lactate_day1 ~ weight_admit + insurance + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)

```


```{r}
model <- glm(mech_vent_overall ~ admission_age + gender + race_group + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)
```


```{r}
model <- glm(rrt_overall ~ admission_age + gender + race_group + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)
```


```{r}
model <- glm(vasopressor_overall ~ admission_age + gender + race_group + language + weight_admit + insurance + adm_type + SOFA + charlson_comorbidity_index + resp_rate_mean +	mbp_mean +	heart_rate_mean +	temperature_mean +	spo2_mean, data = mydata)
summary(model)
sigma(model)
```




